---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README_files/README-",
  fig.width=8,
  fig.height=8,
  fig.units="in"
)
```

# grcdr

This package provides a collection of custom `ggplot2` extensions - `Geoms`, `Stats`, `Coords`, `Themes`, etc. The name `grcdr` is a contraction of *Graphs with Code and Data in R* and is a reference to the [*Graphs with Code and Data*](http://graphswithcodeanddata.xyz/) blog, to which this package is loosely related.

## Installation

You can install `grcdr` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bdilday/grcdr")
```

## Examples

The below will use `dplyr` and `ggplot2`

```{r load2, message=FALSE}
library(grcdr)
## load basics
library(dplyr)
library(ggplot2)

## set the theme
ggplot2::theme_set(theme_minimal(base_size = 14))
```


## geom_excursion

`geom_excursion` plots running quantities as connected scatter plots. It requires an `x` and `y` aesthetic and also a `t` aesthetic to give the ordering (i.e. "time")

#### MLB runs scored and allowed 

Here's some example data provided with the package that gives team-level stats for 4 MLB teams (Cleveland 1999, New York Yankees 1998, Houston Astros 2017, Milwaukee Brewers 1982). 

``` {r runs1}
# load some example data from the package
teams_df = read.csv(system.file("extdata/team_stats.csv", package = "grcdr"),
                    stringsAsFactors = FALSE)
teams_df$game_date = as.Date(teams_df$game_date)
head(teams_df, 2)
```

We can plot runs scored (`b_r`) on the x-axis and runs allowed (`p_r`) on the y-axis. The time coordinate is the season game number.

``` {r}
base_plot = teams_df %>% 
  ggplot() + labs(x="runs scored", y="runs allowed")
```

By default the data aren't averaged (the run length is 1)

``` {r}
p = base_plot + 
  geom_excursion(aes(x=b_r, y=p_r, t=season_game_number)) + 
  facet_wrap(~k)
print(p)
```

The `run_length` parameter controls how many items get summed. The following uses `run_length = 10`

``` {r}
p = base_plot + 
  geom_excursion(aes(x=b_r, y=p_r, t=season_game_number), run_length = 10) + 
  facet_wrap(~k)
print(p)
```

##### averages instead of sums

If you give `x_weight` or `y_weight` aesthetics then weighted averages are computed instead of sums. Passing `_weight = 1` therefore results in straight averages.

``` {r}
p = base_plot + 
  geom_excursion(aes(x=b_r, y=p_r, t=season_game_number, x_weight=1, y_weight=1), run_length = 10) + 
  facet_wrap(~k)
print(p)
```

##### time ordering

Additionally, the time ordering can be changed

``` {r}
set.seed(101)
random_idx = sample(1:nrow(teams_df), nrow(teams_df))

# order by random
p = teams_df %>% cbind.data.frame(random_idx=random_idx) %>%  
  ggplot() + labs(x="runs scored", y="runs allowed") + 
  geom_excursion(aes(x=b_r, y=p_r, t=random_idx, 
                     x_weight=1, y_weight=1), 
                 run_length = 10) + 
  facet_wrap(~k)
print(p)
```


## stat_run

`stat_run` is a lower-level utility than `geom_excursion`. It can change the `geom` (from `path`) and also can plot running line charts in addition to connected scatterplots.

##### smoothed line charts

Here I set the y aesthetic but not x, which is effectively a line chart.

``` {r}
p = teams_df %>% ggplot() +
  stat_run(aes(y=b_r, t=season_game_number)) + 
  facet_wrap(~k) + labs(x="game number", y="runs scored")
print(p)
```


However, unlike a traditional line chart, I can average or sum the y variable

``` {r}
p = teams_df %>% ggplot() +
  stat_run(aes(y=b_r, t=season_game_number), run_length = 10) + 
  facet_wrap(~k) + labs(x="game number", y="runs scored")
print(p)
```

The default is to generate running totals by using the cumulative sum function and taking differences. A different cumuklative aggregation function can be specified, however. Note that following is for illustration and that the run_fun doesn't correspond to any particularly useful quantity(that I'm aware of, anyway).

``` {r}
fun_with_cumsum_fun = function(x) {
  cumsum(x * (x - 1) * sin(x / 5 * pi))
}

p = teams_df %>% ggplot() +
  stat_run(aes(y=p_r, t=season_game_number), 
           run_length = 10,
           y_run_fun = fun_with_cumsum_fun) + 
  facet_wrap(~k) + labs(x="game number", y="runs scored")
print(p)
```

##### reproducing a geom_excursion

Setting the `x` aesthetic reproduces a `geom_excursion`

``` {r}
p = teams_df %>% ggplot() +
  stat_run(aes(x=b_r, y=p_r, t=season_game_number), run_length = 10) + 
  facet_wrap(~k) + labs(x="runs scored", y="runs allowed")
print(p)
```

##### fill missing values

If the time coordinate is missing values, we can fill in the corresponding x and y.

``` {r}
censored_df = teams_df %>% filter(season_game_number < 40 | season_game_number >60) 
p = censored_df %>% 
  ggplot() + 
  stat_run(aes(y=b_r, t=season_game_number, x=season_game_number)) + facet_wrap(~k)
print(p)

p = censored_df %>% 
  ggplot() + 
  stat_run(aes(y=b_r, t=season_game_number, x=season_game_number), 
           run_length = 10)  + 
  facet_wrap(~k)
print(p)

p = censored_df %>% 
  ggplot() + 
  stat_run(aes(y=b_r, t=season_game_number, x=season_game_number), 
           run_length = 10, y_run_fill_value = 20, x_run_fill_value = 50) + 
   facet_wrap(~k)
print(p)
```

##### other geoms

The `geom_excursion` layer forces a `path` `Geom`, but the lower-level `stat_run` layer can change the `Geom`. For example it can use `polygon` (although unclear what the interpretation is)

``` {r}
p = teams_df %>% 
  ggplot() + 
  stat_run(aes(x=b_r, y=p_r, t=game_date), geom='polygon') + 
  facet_wrap(~k)
print(p)
```

## StatMDS

This stat applies dimensionality reduction using multi-dimensional scaling. As of this writing the available algorithms are principal components analysis (`pca`) or t-distributed stohcastic neighbor embedding (`tsne`). The variables to use in the dimensionality reduction are passed in the aesthetics `x#` where `#` is an arbitrary integer. The default `geom` is `GeomPoint`.

#### example data

``` {r}
set.seed(101)
df1 = data.frame(x1 = rnorm(100))
for (i in 2:10) {
  k = sprintf("x%d", i)
  df1[,k] = rnorm(100)
}

# now, for the last 25 add a constant to create two well separated groups
df1[75:100, ] = df1[75:100,] + 2
```

#### pca

Use only 2 variables

``` {r}
set.seed(101)
p = df1 %>% ggplot(aes(x1=x1, x2=x2)) + 
  stat_mds(mds_method = "pca")
print(p)
```

Use them all

``` {r}
set.seed(101)
p = df1 %>% ggplot(aes(x1=x1, x2=x2, x3=x3, x4=x4, x5=x5, 
                       x6=x6, x7=x7, x8=x8, x9=x9, x10=x10)) + 
  stat_mds(mds_method = "pca")
print(p)
```

Use them all and label them

``` {r}
set.seed(101)
p = df1 %>% mutate(rn=row_number()) %>% 
  ggplot(aes(x1=x1, x2=x2, x3=x3, x4=x4, x5=x5, 
             x6=x6, x7=x7, x8=x8, x9=x9, x10=x10)) + 
  stat_mds(mds_method = "pca", geom="text", aes(label=rn))
print(p)
```

#### tsne

Apply t-SNE. This requires the `Rtsne` package.

``` {r}
set.seed(101)
p = df1 %>% mutate(rn=row_number()) %>% 
  ggplot(aes(x1=x1, x2=x2, x3=x3, x4=x4, x5=x5, 
             x6=x6, x7=x7, x8=x8, x9=x9, x10=x10)) + 
  stat_mds(mds_method = "tsne", geom="text", aes(label=rn))
print(p)
```

## GeomTailScatter

This geom implements a tail scatter plot. It is inspired by the [xenographics project](https://xeno.graphics/). The `x` and `y` aesthetics are points in a two-d plane. Subsequent variables are passed in aesthetics named `x#` where x is an arbitrary integer. They do not need to start at `1`, however, the order will be interpreted lexigraphically. The `x#` variables are mapped to lines extending at an angle of `-(15 + 30 * i)` degrees. This means that variables trying to use 12 or more variables in additon to `x` and `y` is not supported at this time and will result in lines that overlap. 

### geom_tailscatter example

Some simulated data

``` {r}
set.seed(101)
df1 = data.frame(x1 = rnorm(100), x2 = rnorm(100))
df1$x3 = with(df1, x1**2 + abs(x2))
df1$x4 = 100 * df1$x1 ** 2

# make a categorical var
df1$g = factor(sample(c(0,1), 100, replace = TRUE))
```

Plot with `geom_tailscatter`

``` {r}
p = df1 %>% 
  ggplot(aes(x=x1, y=x2, x3=x3, x4=x4)) + 
  geom_tailscatter(size=2)
print(p)
```

The parameter `tail_scale` controls the length of the tail lines

``` {r}
p = df1 %>% 
  ggplot(aes(x=x1, y=x2, x3=x3, x4=x4)) + 
  geom_tailscatter(size=2, tail_scale = 0.5)
print(p)
```

Color by group

``` {r}
p = df1 %>% 
  ggplot(aes(x=x1, y=x2, x3=x3, x4=x4, color=g)) + 
  geom_tailscatter(size=2)
print(p)
```


